R Notebook

### 2016-07-14 finalwriteup/trial_code_eq.rmd
finished reviewing lit review

---
title: "draft 1"
author: "Amy Cook"
date: "Tuesday, July 07, 2015"
output:
  word_document:
    reference_docx: mystyles.docx
  html_document: default
bibliography: Library.bib
---

@@

# Keywords

# Abstract

* Abstracts allow readers to quickly decide whether the research is relevant to their own interests
* get at your main argument quickly
* explain the purpose, goal methods, and results

* why did you do the study?, what is the problem
performing loss-making jobs for fixed fees is a major problem in the consulting, particularly in the competitive construction industry.

* why is it important?
affects the porfitability of a company which impacts the careers, internal morale, and even viability of a company.

* why should someone read your essay?
learn if a company's passively collected data could help them avoid loss making jobs or tactfully enforce strictier contracts.

* what is your main claim?
improve profitability by 9% by avoiding ~4% of projects is a possible way to interpret the agorithm output. other ways are possible.

* discuss variables and your approaches
variables: main employee, % time by each type of employee, team size, client charactersitcs, project characteristics (size, type of work, length)
approaches: complex data with mnay variables, thousands of data points. tried regression as well as machine learning algs: randomf roest and boosted trees, bayesian networks, naive bayes.

* describe evidence to support your claim
hundred of cross validated trials using the 5 methods, as well as 5 blending methods. obtained a 95% confidence interval for each method and simple logistic regression of three models outperformed the others, closely followed by simple averaging. the 95% confidence interval across all cross validated trials was __ to __ profit improvement.

* overview of most important sources

* Results: answer , was hypothesis supported
hypothesis was supprted: CAN improve bottom line of business by mobilising their passively collected data into mL algorhtms.

* implications of your work, meaning
more work to be done on testing the user interface and user ailities further. could assist managers in making better decisions about which projects they should reject or develop a more careful contract structure to improve profits.

# Table of Contents

# List of Figures

# List of Tables

# List of Abbreviations

**ANOVA** - Analysis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.

**AUC** - Area under the curve. In this project, the area refers to the area under an ROC curve. Areas greater than 0.5 indicate predictions performing better than random chance while an area of 1 indicates perfect predictions.

**B2B** - Business to Business. Commerce transactions between two businesses.

**BIM** - Building Information Modelling. A digital representation of physical characteristics of an object or structure.

**CLV** - Customer Lifetime Value. Prediction of the profit a customer will bring over the course of a business' relationship with the customer.

**CBR** - Case based reasoning. A method of estimating by using the results of similar cases.

**CRM** - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.

**DA** - Discriminant analysis. Statistical analysis to predict a categorical response variable.

**DAG** - Directed acyclic graph. A graph consisting of vertices connected by edges, where each edge directs a vertex to another vertex. The arrangement of the edges is such that a vertex cannot follow a path that loops back to itself. 
**ESS** - Error sum of squares. The sum of squared errors from a set of predictions.

**FP** - False positive. A case marked 'negative' in a dataset is incorrectly predicted 'positive' by an algorithm.

**FPR** - False positive rate. The proportion of positives that were incorrectly identified (false positives).

**FN** - False negative. A case marked 'positive' in a dataset is incorrectly predicted 'negative' by an algorithm.

**FWLS** - Feature weighted linear stacking. A linear combination of model results which are interacted with original variables. This allows certain models to be weighted heavier for certain values of the original features.

**IT** - Information technology. The use of computers for storing, retrieving and processing information.

**IQR** - Interquartile range. A measurement of spread or variability for a continuous variable. When the data is ordered numerically and divided into four equally sized portions, the values that divide the portions are called quartiles (first, second, third and fourth quartiles). The first quartile has the highest value of the lowest quarter. The interquartile range is the third quartile value minus the first quartile.

**RMSE** - Root mean squared error. The square root of the mean of the ESS.

**ROC** - receiver Operating Characteristic. An ROC curve plots the performance of a binary classification algorithm its scoring threshold is varied.

**TN** - True negative. A case marked 'negative' in a dataset is correctly predicted 'negative' by an algorithm.

**TP** - True positive. The case where a case marked 'positive' in a dataset is correctly predicted 'positive' by an algorithm. Generally, 'positive' is assigned to categories for which the use case needs to be alerted.

**TPR** - True positive rate. The proportion of positives that were correctly identified (true positives).

# Acknowledgements

# Publications

Cook, A., Wu, P., & Mengersen, K. (2015, September). Machine Learning and Visual Analytics for Consulting Business Decision Support. In Big Data Visual Analytics (BDVA), 2015 (pp. 1-2). IEEE.

# Statement of Original Authorship

The work contained in this thesis has not been previously submitted to meet requirements for an award at this or any other higher education institution. To the best of my knowledge and belief, the thesis contains no material previously published or written by another person except where due reference is made.

Signed .................
Date....................

# Chapter 1: Introduction

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of this sort every day - estimating how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the task is new and difficult. These errors are trivial in everyday activities, but for consulting businesses that solve complex problems, similar time estimation errors determine their financial well-being.

Consulting businesses give expert advice to other professionals in exchange for a fee based on the amount of time their consultanting staff spend on the project. The deliverable 'expert advice' is generally in the form of a document or a package of documents. A document could range from a 5 page report written by a single person to a sequence of hundreds of drawings assembled by a team of 10 over many years. Despite the range of possible deliverables, before any project commences, the client and consultant must agree on a fee, or at least a fee structure. The fee structure can be as creative as the engaged parties wish, but in fact tend to fit a few common models. A couple of simple examples are:

1. The client agrees to pay for the consultants' time by the hour until the task is complete (referred to as hourly rate) or 
2. the consultant offers a fixed fee to complete the project in full, regardless of time spent (referred to as fixed price or fixed fee).

In each of these contract models, financial risk is assigned to one of the parties. In model 1, the client must pay for an unknown quantity of consultants' billable hours and in model 2 the consultants must complete the project for a fixed price while not knowing how much of their own time will be spent.

This thesis focuses on the risk taken by a consulting company when offering a fixed fee. Clients commonly seek fixed price quotes from several consultants before selecting one or negotiating further. Quoted fees must be competitive.

The way a consulting manager calculates a fixed price varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project they are quoting and, after reviewing the project details, can use a combination of intuition, comparison to past projects, and rules of thumb (such as a set fraction of the entire client budget for a large project). Such qualitative methods can result in profit margins which ultimately depend on the skill (or luck) the manager has in predicting the amount of time a project demands.

This thesis focuses on a single case study consulting company in the construction industry that has persevered through many fixed price projects that resulted in losses. A source of insight into the loss-making projects was identified - twelve years of passively collected data that described each project in terms of their clients, invoice history, employee hours and technical details. Was it possible to find trends in what makes a project loss-making and can these characteristics be identified before a fixed price is offered? Can these trends be modelled into a predictive algorithm, and if so, how could the outcomes be integrated into business decisions? Finally, if the predictive outcomes were integrated into decision-making, would they deinitely improve the business' bottom line? These questions are explored in depth and answered over the course of this dissertation.

Due to the commercially sensitive nature of the data used, the source is not disclosed. Variables have been de-identified to ensure confidentiality is maintained.

# Chapter 2: Problem Description

“It’s tough to make predictions, especially about the future” - Source Unknown

In this chapter the research motivations are explained along with the goals of the project and the specific research hypotheses. Finally the contributions of this body of work are outlined along with the structure of the thesis.

## Research Motivation

**Problem Description:**

Project managers across many industries have always struggled with forecasting project costs. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by an average of 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% [@Flyvbjerg2007]. A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget [@Flyvbjerg2011]. Another review of 6 IT project surveys between 1984 and 1994 by @Molokken2003 revealed that 60-80% of IT projects encountered effort and/or schedule overruns, where the average overrun was by 30 to 40%. These statistics describe difficult conditions that can result in job losses, business bankruptcies, and government budget blow outs. A survey from @Moores1992 asked software managers whether they saw effort estimation as a problem, and 91% responded 'yes'.

Smaller consulting companies in competitive industries, such as the construction industry, experience similar difficulties in forecasting their project costs. They must offer appealing fixed prices in order to remain competitive and win projects. However, when these projects go over budget it negatively impacts the business in many ways. Employee morale deteriorates and employees may view themselves incompetent if they are responsible for the project's delivery. Employees also experience elevated stress attempting to complete underestimated projects within a disappearing budget and may compromise quality of work in exchange for speed. This is unfortunate if the only issue with the project was an under-estimated fee. Unprofitable projects also limit a business financially, hindering investment into marketing, training, and business development. The financial risk taken on fixed price projects may also discourage businesses from taking other, more calculated risks, such as expanding the business into a new area. Clearly, improving the prediction of project profitability would improve business growth and staff morale.

**A Wicked Problem**

The challenge for consultants to reduce budget blowouts, in terms of hours spent, is not easily solved. Complex projects always differ from one another - even similar projects may end up demanding significantly different amounts of time. Although this is known, a fixed price is often demanded by clients before a project is awarded. If an hourly rate fee structure is proposed, clients in many industries can simply turn to another consultant willing to take the financial risks involved with fixed prices **(...ref...)**. Two strategies to guard against fixed price budget blowouts are detailed below along with their limitations.

A compromise between the hourly rate and fixed price structure is to track employee hours, and stop work once the fixed consulting fee has been exhausted. Then, negotiate further fees, or variations, with the client. This can either be agreed upon with the client beforehand or the consultant's internal strategy. If this strategy is proposed to the client up front, it is likely the client would again turn to another consultant who is willing to offer a fixed price. If this approach is the consultant's internal strategy, their financial risk may satisfactorily be reduced. However, the behaviour may cause friction between the client and the consultant, therefore reducing the potential for repeat business. In the event of hours inflating beyond expectation, consultants may choose to wear losses or endure marginable profits to preserve their client relationship. They may reason it is in their best interest to maintain a reliable, trustworthy reputation in their industry above creating financial friction.

Another strategy for managing this risk is by stipulating a specific one-off variation opportunity in the contract. This applies to consulting work that falls under the client's larger project budget. The variation would specify a percentage fee of the overall project cost that can overrule the original fixed price. The purpose of this rule is to guard against the case where the project creeps in size, and the consultant finds themselves gradually undertaking more work and more work for the same price even though the client was happy to spend more on the overall project. This may account for some or in the best case, all, of their overspent time, although again a savvy client may turn to other consultants that agree to omit variation opportunities.

Clearly, there is not a straightforward contractual solution to reducing the financial risk taken on by consultant's who offer fixed prices in complex projects. In many industries, the reality of fixed price fee structures are unavoidable and businesses conform to survive.

**Current Fixed Price Estimation Methods:**

It has been established that estimating fixed price fees in consulting is a difficult task with unfortunate consequences if the price is estimated too conservatively. Thie project aims to mobilise a consultant's historic data describing profitable and loss-making jobs in order to improve their fixed price setting practices. But what is the current method for estimating fixed prices or project costs? Is past data currently used in other cost estimation practises and to what degree of success?

Unfortunately, limited research is available that documents the methods industry uses for complex-projects. However, two surveys are available: one from 2000 that covers construction project cost estimation, the other from 1992 and focuses on IT project effort estimation (which can be likened to fixed price estimation for consultants). The construction cost estimation study by @Akintoye2000 surveyed 84 UK construction contractors, ranging from small to medium to large, about their estimating practices. They found that the main method was breaking the project into detailed parts and summing up the cost of each item. The next two most popular methods were 'comparison with similar projects based on documented facts', and 'comparison with similar projects based on personal experience'. These can all be classified as experience based models [@Akintoye2000].

A survey by @Moores1992 of 54 software developing companies found that detailed project planning tools were used by most companies as opposed to cost estimation tools, suggesting that projects were priced based on an analysis of a detailed breakdown of tasks within a project. In the same survey, 91% of software companies cited cost estimation as a problem.

Both surveys indicated cost and effort estimation were performed via a detailed breakdown of the tasks required. Analysis of past project data was alluded to (comparison of similar projects) but a rigorous data analysis process was not common. Therefore it is worth investigating if a predictive model based on past data could improve the fixed price setting practice.

**Cost Estimation and Human Nature**

Although the surveys indicated industry did not analyse past project cost data, the idea of using this data (in the same way) has been studied before. Research exists that tested people's estimation accuracy with and without comparitive historic data with interesting results. This is summarised below along with the resulting data-driven models.

@Lovallo2003 delved into the psychology behind why executives so often severely underestimate costs of larger projects such as manufacturing plant construction, mergers and acquisitions, large infrastructure and software development. Their theory stemmed from Kahneman's work on decision-making that won him the Nobel Prize for economics in 2002. His research argued that a person's natural optimistic view of their own skills leads to consistent underestimation of the time and risks involved in a project. A manager optimistically sees challenges in a project as something that can be overcome by the team's high skill level, and downplays or ignores the risk of problems that are out of the team's control. This is why it does not matter if the project is broken down to the highest level of detail for cost prediction, all complex projects are at risk of encountering a multitude of problems that the manager could never foresee. Each problem has a low chance of occurring, but in combination the risk is much greater [@Lovallo2003].

@Lovallo2003 call the practice of analysing a project based on project details without considering unforseen risks the 'inside view'. A survey by @Moores1992 aligned with this theory and found the most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%). This contrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project should be ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommended was to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecasted cost made by detailed analysis (the 'inside view') [@Flyvbjerg2011]. As demonstrated in the surveys, @Flyvbjerg2011 found that managers rarely analysed data from similar projects even though it could significantly improve cost predictions. This discovery led the team to formally introduce 'outside view' methods into some institutions.

The implementation of reference class forecasting began in project management for the first time in 2004. It was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, and Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecasted cost [@Flyvbjerg2011]. An example reading from the chart is as follows: "If the chance of a cost overrun of 20% is acceptable, the estimated cost should be increased by 55% from original calculations.". This builds past experiences of cost overruns into the initial cost estimation.

![This model was implemented by the British government for a rail project that is still under completion [@Flyvbjerg2011]](images/lit_review/UKrail.JPG)

The visualisation presents a powerful communication tool for influencing decision makers and improving forecast cost accuracy, however, access to credible data for a sufficient number of projects can be a challenge [@Flyvbjerg2011]. Overall, the optimism bias theory credibly helps to explain the persistant problem of project cost underestimation.

### Further Reference Class Forecasting Models and Industry Uptake

Although @Flyvbjerg2011's reference class forecasting theory and the psychology behind it was first published in 2003, predicting project costs using previous data has been researched for many preceding years and continues to this day. In this section, a brief overview of other reference class forecasting models is presented (they are reviewed in greater detail in the literature review) followed by an analysis of why industry has not yet adopted them.

Most cost estimation or reference forecasting analytical models focused on software development and construction projects, along with some work on film projects. These analyses generally used information from 15 to 20 large projects and the dataset summarised projects from around the world with different clients and teams. Applied mathematical methods varied from statistical Regressions to deep-learning Neural Networks [@Love2005][@Kim2004]. Most studies reported that their predictive models performed well and that Neural Networks were generally more accurate than Linear Regression.

#### Industry Uptake of Existing Models

Despite the predictive success in statistical and machine learning cost estimation models as well as the revealing outcomes from 'inside' and 'outside' view studies, reference class forecasting there has been a lack of industry uptake. Two reports reviewing cost estimation in the software industry investigated reasons why data-driven forecasting models have not been translated to industry. The key causes were:

* A lack of a decision framework to support manager's use of the 'outside view' model [@Moores1992]
* A failure in marketing of the potential for the models to improve business outcomes [@Moores1992]
* The products of the research were not comprehensive enough to be used in industry [@Jorgensen2007]
* Expert judgment was not mobilised enough in the process of developing the tools and the resulting product felt disconnected to actual requirements [@Jorgensen2007].

Reports on other industries were not available, however the insights are general and may explain the lack of adoption in other industries (such as the construction industry).

In the software industry it is also possible the problem of fixed price contract blowouts has been solved by an alternative contract structure which has made cost estimation models unnecessary. In contrast to construction, IT projects are a relatively new practice and the industry has had success in trialing alternative contractural arangements. Through experimentation, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted and revised based on frequent progress review points that assess unexpected problems. The flexible upfront arrangement with clients has often been found to forge better long term relationships [Cockburn2001]. This agile methodology suits the nature of software development work, where curtailing the scope of work is relatively straightforward. Also, since the industry is more modern, the new methodology was able to sweep across the field and could be proven to deliver better budget and product outcomes to clients [@Rasmusson2010].

Unfortunately, in other industries such as the construction industry, infrastructure and building projects are steeped in a traditional process, where contractual norms have developed over a much longer time period [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way projects are estimated, and contractually bound. Furthermore, the industry would struggle to take on an agile approach as the scope cannot be easily reduced as the project unfolds - the project must result in construction and building certification which implies a high level of legal responsibility and due diligence.

This project aims to assist a company in the construction industry in managing problematic fixed price projects, since fixed price contracts are unavoidable. The blocks preventing industry adoption, such as providing a clear decision framework and incorporating expert judgment will be carefully addressed.

### Case Study

The case study company is an engineering consulting company in the construction industry. In their profession, labour is the chief cost and the long traditional history of this industry mandates fixed price projects as the norm. In recent years, the size of losses from unprofitable projects equalled 25-30% of their profits from profitable projects. This demonstrates the magnitude of their cost estimation problem and the substantial room for improvement that a data-driven model may facilitate.

Cost estimation is currently performed in the case study company by first carefully reviewing preliminary drawings where time costs for each task are estimated and summed by an experienced manager (the 'inside view'). This is often cross-checked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

* lack of time to interrogate past project data
* clunky data availability offered by their database software
* lack of awareness of similar projects performed by other managers

A Customer Relationship Management software package (CRM) is currently employed to collect and store project data. A CRM is a popular type of software used by businesses to record client and project details as well as employee timesheet records. In the case study business, the CRM is available to all employees over the company intranet, and each employee completes daily timesheets alotting their hours to certain projects. Additionally, technical information is recorded against each project as well and the CRM data stood as an untapped source of information within the case study organisation. It stores a rich variety of data including:

* employee timesheet hours with dates
* other project costs (taxis, printing)
* client information/characteristics
* client identification code
* invoiced amounts for each project and dates
* employee costs
* employee charge out rates
* project description

The CRM software readily provides simple output statistics, however, analytical capabilities were limited to:

* simple scatter plots and bar charts of the raw data
* summaries such as overall hours spent vs. the invoiced amount for each project

The company had a wealth of data but limited means to extract insight. This study exploits the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model to improve cost estimation. Thousands of past projects are available for analysis, as opposed to previous studies that used on average 15-20 cases and at most 300 [@Finnie1997][@Pai2013][@Shin2015]. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on internal company data. Managers could interrogate how the model outputs were calculated and relate to the actual projects used in the model or speak to a colleague that was involved. This may give the presented 'outside view' a stronger meaning and more potential to influence the final fee proposal.

### Case Study Limitations

The case study provides an opportunity to test whether passively collected CRM data can be mobilised into a predictive model to improve fixed price cost estimation. However, there are some limitations in the study including the size of the study, missing data, and errors in data entry. This section addresses how the limitations can be addressed either in future work or within the research framework.

The most obvious limitation is that the work focuses on a single company in a specific industry, which limits the application of conclusions to a broader range of companies. This can be overcome by first determining the optimum method and variables for this case and then testing them on future case study companies.

Another limitation is the type of unavailable data. Information on overall costs of construction projects was scarce. This is relevant because managers sometimes price jobs based on the budget for the entire project, of which the consultant is a minor part. For example, engineers may moderate their fee by 1% of the entire project cost. In the future, it would be beneficial for managers to record the expected total project cost at the beginning of a project versus the concluding completed cost. Another piece of data not recorded was the external factors influencing the fixed price quote including known competition from similar firms.

The database also excludes detailed technical information about each project. Such details are found in the preliminary drawings describing the project (square meterage, number of storeys), as well as project summary documents provided by the client to enable a quote. However, this level of information actually contradicts the purpose of the 'outside view', the intention of which is to present similar projects without being over-influenced by finer details. Given the lack of detailed information describing the physical size of projects, it was not reasonable for an algorithm to predict the numeric fee. Instead, it was more logical to predict the profitability of a project i.e. return per dollar. This measure represents the ratio of the amount invoiced to business costs, and can easily be compared between projects of different sizes.

Finally, all CRM data is entered by employees manually. Because the data is input by people, it is susceptible to some error. Many errors were detectable during data cleaning, however it is possible that a low number of undetectable human errors exist in the data and marginally influence predictions.

### Research Motivation Summary

Fixed price cost estimation for complex projects is a difficult task and when performed incorrectly, cause substantial financial losses to businesses and governments. Studies have explored human nature and our tendency to optimistically assess our capabilities with respect to a given task. In the case of cost estimation, the optimism can result in negative business consequences affecting employee morale and the capacity of a company to flourish and grow. Over the past few decades, significant research has been dedicated to creating predictive models that present the 'outside view' of a project. This was done by statistically comparing a new project to a collection of similar projects and their characteristics. It has been shown that these models improve cost estimation accuracy, however industry has not adopted them. Managers tend to predict costs based on intricate details of a project in favour of reference class forecasting. This may be due to poor marketing of the cost estimation tools, a lack of a decision framework in relation to the tools, and a lack of collaboration with industry during model development.

## Thesis aims

For the reasons explained above, the aim of this research is as follows:

**General Aim** Use statistical techniques to model the profitability of projects for consulting businesses using their internal CRM data. Research will focus on a case study Engineering consulting company that offers their expert advice (in the currency of time) to business clients. 
The project outcomes are intended to assist the business in predicting project profitability before engagement. Several statistical and machine learning techniques were tested, compared and refined.

**Hypothesis 1**
A statistical or machine learning model based on historical project data can predict the profitability of a new project.

**Hypothesis 2**
The predictive model built from Hypothesis 1 can be shown to have a positive impact on the bottom line of the case study business.

## Thesis Contributions

In comparison to previous studies, this project advances the body of work on cost estimation in three ways: the dataset is larger than other studies and sourced from a single company's CRM database, ensemble tree methods and model blending was tested, and the prediction results were integrated into a decision framework for the business. These points are outlined in more detail below.

In past research, cost estimation datasets generally consisted of dozens of projects from a range of a companys and even countries. This thesis case study only had access to the company's internal data but this contained over 2,000 past jobs - far more than the usual cost estimation study and with many more variables. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry which differs from other studies in the construction industry that estimate actual construction costs. Consulting in the construction industry can be likened more to software development effort estimation although the nature of the work differs considerably.

In addition to addressing cost estimation with a different kind of dataset, ensemble tree methods are applied which have been minimally tested on this problem. To date, Linear Regression, Neural Networks, Case Based Reasoning, and Support Vector Machines (SVM's) have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks in some cases and generally outperform Linear Regression (**ref**). Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots that provide insight into the model's calculations. This contrasts Neural Networks and SVM's which are 'black-box' predictors. Providing insight to a user is a major benefit as past research has demonstrated issues with user trust and uptake of models, even if the models predicted well. It is anticipated that using internal data will also improve trust and create tangible meaning for the user. Blending multiple machine learning and statistical models has also not yet been applied to the cost estimation problem and may improve predictions.

Finally, this study propels the predictive model one step further than other studies by analysing bottom-line profit improvements that could result from the research. A few possibile ways to integrate model results into business decision-making scenarios are presented, and where possible, final profit increases are evaluated. This kind of thought experiment is designed to present a clear case for industry when evaluating if and how to adopt cost estimation models.

## Thesis Structure

The next chapter, Chapter 3 is a thorough review of the literature to date that relates to this thesis. This covers work solving the cost-estimation problem, studies on statistical and machine-learning models in business applications, and a review of effective statistical and machine learning methods.

Chapter 4 describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicting profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision making scenarios. Further applications were also considered.

The research outlined in the method were divided into four stages, each of which was assigned a chapter containing the results and discussion of that portion:

* Chapter 5: Variable Selection
* Chapter 6: Regression Models
* Chapter 7: Binary Classification Models
* Chapter 8: Model blending,
* Chapter 9: Extended Analysis including overall profit analysis, research limitations and future work

The results section in each of the chapters listed above presents key charts and tables that progressively answer the research question: can a statistical or machine learning model based on a company's internal historical project data predict the profitability of a new project and have a positive impact on the bottom line of the business? Then, a discussion finalises each chapter, where the results are examined more deeply and analysed in a broader context. The nature of the results are also debated in terms of how surprising they were, their impact on the research hypotheses and limitations as well as suggested future work. The final chapter, the Conclusion, summarises the findings of the project, how the research answered the hypotheses and how the work could be viably applied in industry.

# Chapter 3: Literature Review

"If I have seen further than others, it is by standing upon the shoulders of giants." - Isaac Newton

Researching outcomes of comparable past studies may impart how to best approach the problem and reveal shortcomings that can be improved upon. This chapter provides an overview of the literature describing early and current cost estimation methods. The most prominent machine learning and statistical methods shall then be reviewed followed by a summary of papers on more general business applications that have tried a wider variety of methods. Gaps in past research will also be highlighted.

## Project Cost Estimation Methods

The practice of estimating fixed fees is usually represented by the terms 'cost estimation' or 'effort estimation' in the literature. The bulk of research to date has been performed with project data in either the construction industry or software development. The different methods and outcomes in each of these industries is reviewed below. Finally a gap in the literature will be highlighted which this research addresses.

### Cost Estimation in the Construction Industry

Research in the construction industry has primarily focused on predicting the final building cost of construction. Some studies reported surveyed cost estimation practices while others tested new methods such as building information modelling (BIM) analyisis, case based reasoning, and predictive algorithms. The cost estimation methods of detailed analysis (the 'inside view'), Case Based Reasoning, and predictive algorithms are reviewed in this section.

#### Detailed Analysis in the Construction Industry

Detailed analysis refers to the process of an engineer or builder in the construction team carefully reviewing construction drawings to sum the cost of materials, labour, machinery hire, overheads, and profit etc. [@Akintoye2000]. As referenced in the Research Motivation, @Akintoye2000's study found that this was the most popular method amongst contractors.

@Elfaki2014 discusses how much detailed cost estimates for construction projects can vary from estimator (engineer) to estimator. This contributes to the lack of accuracy in predicting final costs of a project. He argues that an engineer's expertise is not documented or measured in any way and therefore their expertise and, in turn, their estimate is prone to subjectivity. @Shane2009 theorises that final cost can be subject to so many different unpredictable parameters, such as weather, unexpected ground conditions, duration or sub-contractor issues, that it is almost impossible to achieve an accurate cost prediction manually.

Groundbreaking methods into calculating construction costs by @Ma2014 mobilise data from building information modeling (BIM) models, although this is still technically a detailed analysis of the costs. The idea is to quantify the cost of building a structure directly from a three dimensional BIM model created by the design consultants (engineers and architects). @Ma2014's first trial was to automate the cost calculation of a reinforced concrete structure. He programmed his algorithm to intelligently establish construction techniques for each element as this is not provided by architectural or engineering models. From the construction technique and material, a cost was derived for each element. This was successful and @Ma2014 aims to further his work so that more construction information can be intelligently obtained from BIM models. A system like @Ma2014's would definitely accelerate the cost estimation process and reduce human error, however it is still an 'inside view' and could still deliver innaccurate estimates similar to detailed analysis by a person. The detailed view does not 'step back' and take into account set backs in construction that may have affected past similar projects. There is potential for this innovation to be combined with an 'outside view' system.

Detailed analysis remains the most prevalent method for construction cost estimation despite the industry having a long history of projects running over time and budget using the same technique [@Shane2009]. Next, the method of detailed sums shall be compared to results from parametric algorithms and case based reasoning ('outside view').

#### Statistical and Machine Learning Models in the Construction Industry

Dozens of studies have built machine learning and statistical models to predict construction costs. This section will first review the previous studies and most accurate algorithms and finally summarise industry uptake of the methods.

@Elfaki2014 completed a review of cost estimation research from 2004 to 2014. They found that artificial Neural Networks and SVM's were the most common machine learning techniques. He highlighted that they dealt well with uncertainty, however lack technical justification to the decision maker (known as a black box predictor). Neural networks are also time consuming to train, and must be re-trained and re-tested with each additional piece of data [@Kim2004]. Nevertheless, Neural Networks and SVM's received significant attention in the 1990's for their ability to accurately predict construction costs with limited detailed information [@Shin2015][@Kim2004].

@Shin2015 pioneered the application of Boosted Trees to cost estimation in construction projects. This is surprisingly late given the rapid uptake of Boosted Trees over the past decade. In their study, data from 234 school building construction projects in Korea were used. Boosted trees were compared to Neural Networks and were found to predict costs slightly more accurately than Neural Networks, but not statstically significantly (p value < 0.05).

A notable amount of literature tested the predictive power of Multiple Linear Regression in the construction cost estimation problem. Often, Linear Regression was the only model assessed, without comparison to other methods such as Neural Networks, which first started appearing in literature in the 1990's [@Kim2004]. However, even post-1990, many studies focused on Linear Regression only. This may be because the technique is straightforward, easy to use and widely available in statistical packages [@Chan2005]. Some studies showed that Neural Networks outperform Regression, however other studies established they are approximately equal [@Kim2004][@Attalla2003]. @Dissanayaka1999 suggested that Regression models could be performed first to determine variable importance and condense the number of variables, then Neural Networks could be used for better estimation. Boosted Trees may perform as well as Neural Networks while providing even more insight than Regression models because it is possible to produce partial dependency plots. These visualise the relationship between each variable and the response variable (cost).

Regression models from different studies tended to use a wide range of variables, each of which influenced the models to varying degrees. This is likely because construction projects are multidisciplinary and involve many parties such as the clients, consultants, contractors and suppliers, [@Chan2005]. @Chan2005's study on 87 building projects in Singapore included special, complex projects. Therefore, variables such as contractor's specialised skills, who the client was (public vs. private), the client level of experience, and the contractor's financial management highly influenced final cost of the project. Other studies found variables such as project complexity, duration, team experience, information availablility, site requirements, and labour climate to contribute most [@Akintoye2000][@Trost2003][@Pinto1988]. As these studies obtained information from questionnaires, the available variables depended on the questions. Despite the differences in qualitative outcomes, the studies all generally reported their mean absolute error or a similar metric from the tested models, and results were generally positive.

Similar result statistics are useful, however there is a gap in the literature rationalising which model industry should adopt, how to adopt it into a business, and how much loss can be reduced. This gap may explain why uptake of these algorithmic models are sparse, as indicated by @Akintoye2000's survey, despite 20 years of prior research. @Akintoye2000 hypothesised this could be due to a lack of knowledge about the techniques, doubts whether these techniques are applicable to the construction industry, and the availability of sound data to ensure confidence.

The literature on construction cost estimation does not acknowledge that dozens of consultants and sub-contractors contribute to a large project. Most consultants, such as engineers and architects, produce their own cost estimates for their effort that contributes to the project. Their contracts are often fixed price, with minimal room for movement, and they wear the risk of over-spending their time [@Harris1999]. Furthermore, the existing studies collected data from many businesses for a single study, which leaves a gap in research exploring a single contractor's history of projects at one time.

#### Case Based Reasoning in the Construction Industry

Case based reasoning (CBR) is a systematic method of expert judgment, where the decision maker manually compares similar projects from his or her experience [@Shepperd1996]. The idea is that in the case of a new project, similar projects are chosen, before generalisations about the data are made. This contrasts to algorithmic models that are built from the entire set of data [@Elfaki2014]. Research into the application of case based reasoning to construction cost estimation began in the 1980's [@Kim2004]. The general procedure involves:

1. Storing a collection of projects with key variable values
2. Once a new project or case arrives, similar cases are retrieved. This can be achieved either algorithmically with a distance function applied to the variables, or manually by users reviewing past cases. The nearest neighbour algorithm has been used to find the distance between cases in previous studies. It is programmed to calculate the Euclidean distance in n-dimensional space between cases, where each variable is a dimension [@Kumar2007].
3. The cost of the new project is estimated by extrapolating characteristics of the similar cases to the new case. This can also be done either algorithmically or manually by an experienced decision maker.

A study by @Kim2004 compared CBR to Neural Networks and Linear Regression in construction cost estimation. For CBR, the ESTEEM software package algorithmically calculated the similarity of variables, weighted the variables using gradient descent, and deduced the cost using the most similar cases (the specific method for this last step was not explained). The discussion compared mean absolute error rates of the three methods and showed that CBR was more accurate than Regression, but less accurate than Neural Networks in estimating cost. Despite Neural Networks outperforming CBR, CBR maintained advantages over Neural Networks which have been referenced in other studies as well.

Advantages of CBR include algorithmic efficiency as well as user engagement. In @Kim2004's study, it was noted that CBR models were simple to update with new data in comparison to Neural Networks, which must be slowly re-trained and re-tuned. New cases also did not need to have every variable complete which can occur often in real, messy data. Neural Networks on the other hand require complete data. Another advantage of CBR is the ability for users to review chosen similar cases and make sense of the prediction, as opposed to Neural Networks [@Kim2004][@Elfaki2014][@Kumar2007]. One disadvantage is that accuracy can be highly dependent on the number of selected cases [@Elfaki2014]. Overall, CBR may provide prediction accuracies between regression and Neural Networks but the results can be justified by the user - a valuable asset.

CBR that is performed manually through personal experience, without the assistance of algorithms, was the second most popular method for construction estimation in @Akintoye2000's survey in the UK. This indicates it is quite an intuitive method that allows decision makers to cross check their detailed cost estimates, although an algorithmic version of CBR has not found success in industry.

>>>>

---
title: "R Notebook"
output:
  word_document: default
  html_notebook: default
---

@@
